Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?

### 2023-03-27 Manuscript/manuscript.tex
Added more to intro

2) and 3)
Other scholarly papers have discussed track and field reaction times but not in the
context of whether the minimum reaction time barrier is too low or whether timing
systems were accurate.  Outside the blog posts from LetsRun.com, there have not been
any published papers regarding Devon Allen's disqualification in particular.  This is not
necessarily surprising given that his disqualification occurred only last August.
Letsrun is a website that is part message board and part news.  The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as "Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and "The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the Letsrun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, statistics such as median reaction time, and a reference to a study
published in cooperation with IAAF in 2009 questioning the validity of the 0.1 
second reaction barrier.  The paper analyzed sprinters reaction times
and the biological processes that went into reacting to the starting gun and
concluded that athletes can react in as low as 0.08 seconds. It even recommended
to IAAF that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommmended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
This could provide a more accurate representation of reaction time, and thus the
possibility of a false disqualification would be lowered \citep{Komi}.
Despite the study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is confusing
as to why IAAF would commission a study only to disregard the results and not
make the necessary changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.

@@

4)
This paper will try to model the data using a 
When determining the most effective model to use, several examples were
explored across the 4 different data sets: a linear model without a year effect,
linear model with year effect, a gamma model without a year effect, a gamma model
with a year effect, a gamma model without a year effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed both
linear models with various distributions to be modeled\citep{Rpkg:lme4}. A GLMM is a type of model
that supports inputs that are from different distributions.  The model that we use is for the
Gamma distribution for the fixed effects part of the model and a random effects part.  In
the model used in the paper, the fixed effect comes in the form of an intercept, and then
there is a random effect which in this case we will call the venue effect.  The
venue effect is responsible for the variation between years and tries to explain
why some years have athletes with faster reaction times than others.

In the code, the 
link parameter under Gamma() was changed to be "log" as logarithmic functions are better 
equipped to deal with high outliers.

@@

Something that is extremely important to note in the Data is the inconsistency of World Athletics 
in terms of how they penalize false starts.  From 2007-2009, the IAAF (the former name for World 
Athletics, and the governing body for the World Track and Field Champions) instituted a rule change
that allowed one warning false start before a sprinter was disqualified \citep{False-Start}  This 
essentially gave sprinters a second chance and reduced the penalty for a false start.  There were 
18 male and 7 female false starts at the 2007 World Championships, 18 male and 7 female starts at 
the 2009 World Championships. Starting in 2011, the rule was scrapped, and the old policy which 
automatically disqualifies runners who false start was reinstituted.  The effect? At the 2011 World
Championship there were 6 male disqualifications and 4 female disqualifications \citep{False-Start}. 
By returning to the harsher policy and cracking down on false starts, World Athletics had reduced the
number of false starts by two thirds. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule.\citep{Pilianidis}

@@

\jy{Describe you data types; give details on sources; introduce notations for
  both types of data
  present exploratory plots.}

@@

Sorted By Model instead:
Linear Model with no year effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Log Likelihood & Int effect est & Int se \\ [0.5ex] 
   \hline\hline
   All Finals & -472.7 & 238.3 & 0.1484 & 0.0019 \\
   \hline
   Older Finals & -450.7 & 227.4 & 0.1495 & 0.0019 \\ 
   \hline
   All Pooled & -1667 & 835.5 & 0.1529 & 0.0014 \\
   \hline
   Older Pooled & -1539 & 771.5 & 0.1544 & 0.0015 \\
   \hline
  \end{tabular}
  \end{center}

Linear model with year effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Year se & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -468.7 & 0.0088 & 237.4 & 0.1481 & 0.0030 & 7.762e-05 & 2.457e-04 \\
   \hline
   Older Finals & -444.1 & 0.0075 & 225.0 & 0.1496 & 0.0028 & 5.691e-05 & 2.470e-04 \\ 
   \hline
   All Pooled & -1713 & 0.131 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\
   \hline
   Older Pooled & -1564 & 0.113 & 785.4 & 0.1562 & 0.0037 & 1.280e-04 & 6.273e-04 \\
   \hline
  \end{tabular}
  \end{center}

Gamma model with year effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Year se & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -491.1 & 0.3388 & 248.5 & 6.8121 & 0.1895 & 0.1148 & 0.0112 \\
   \hline
   Older Finals & -465.1 & 0.2930 & 235.5 & 6.714 & 0.1674 & 0.0859 & 0.0109 \\ 
   \hline
   All Pooled & -1848 & 0.3612 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   Older Pooled & -1688 & 0.3145 & 846.8 & 6.426 & 0.1923 & 0.0989 & 0.0243 \\
   \hline
  \end{tabular}
  \end{center}

Gamma model with year effect and link=log
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Year se & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -491.0 & 0.0504 & 248.5 & -1.914 & 0.0280 & 0.0025 & 0.0112 \\
   \hline
   Older Finals & -465.0 & 0.0439 & 235.5 & -1.901 & 0.0251 & 0.0019 & 0.0109 \\ 
   \hline
   All Pooled & -1848 & 0.0554 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\
   \hline
   Older Pooled & -1687 & 0.0495 & 846.6 & -1.853 & 0.0305 & 0.0024 & 0.0243 \\
   \hline
  \end{tabular}
  \end{center}

@@

Addition of 2022 on Gamma and log model on semifinal and finals data
\begin{center}
  \begin{tabular}{|c | c|}
    \hline
    Dataset & Year Effect Standard Error \\
    1999-2019 & 0.0495 \\
    1999-2022 & 0.0554 \\
    \hline
  \end{tabular}
\end{center}

@@

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a year effect, a batch effect, or a year
and batch effect.  Density plots were created that simulated the probability of a mean
density less than 0.1, which is the current threshold for a reaction time
disqualification.  The density plots are shown below.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.

@@

\section{Further Analysis}
This problem of whether the sensor was faulty can be further explored in
two ways. First we can look at the reaction times of athletes who competed at the
World Track and Field Championship in both 2019 and 2022 and compare the times,
and secondly we can look at how athletes performed compared to how they did at
their national/qualifiers meet.  We would expect that athletes of this caliber,
who are world-class athletes, to not improve significantly from one year to the
next, but that is not necessarily the results that were found.  The R package
"clusrank" was used to perform a clustered Wilcoxon rank sum test.  This was
chosen over a typical t test so as to avoid any assumptions about the distribution
of population means.  The clusrank package allows clusters to be formed, which
in this case are for each athlete.  The data containing each athlete's name,
nationality, and reaction times for the 2019 and 2022 World Championship can be
found in the csv file labeled "2019vs2022".  This data was transformed to "Clusrank2019vs2022"
so that it better fits the clusrank package.  Each athlete was given a non-meaningful
ID (An ID of 1 does not indicate that this athlete is better or worse than any
other athlete).  Additionally, the "Year" column is used to determine what year
the time was registered with "1" indicating 2019 and "2" indicating 2022.

@@

Another important data set that was investigated was the 2022 national competition
data.  In order to send between 1-4 athletes to the World Championships, almost
every athlete had to qualify by proving they are elite at the national level.
Thus, data was compiled on athletes who competed at the 2022 World Track and 
Field Championships on what these athletes also ran at their country's national
championship.  If there was any argument that athletes have improved their reaction
time from 2019 to 2022, it should now be considered nearly null, as the national
competitions took place between June and July 2022 and the World Championships
were in August.  Thus it should be expected that elite athletes such as these
will not get reasonably faster in such a short time span.

The equipment used to measure reaction time cannot be guaranteed
to be the same from national to international competition, but the time span between
meets has been reduced considerably.  Repeating the data collection steps and
again performing a clustered analysis for a Wilcoxon ranked sum test
results in two new z scores and p values.  The "ds" method produced a z score of
3.6069 and p value of 0.000155.  This result is highly significant and shows that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships even for an alpha level of 0.001.

@@

The figure below plots the average reaction time by year for the pooled men's data since
1999.  The circle in the bottom right is for 2022, which is significantly lower than the
data points for the previous three years, and is substially lower than the data points
for 2011 and 2013.  This graph supports the conclusion that the timing system from 2022
had some sort of error that caused the reaction times to be so much lower.  Assuming there
was not an issue with the system, the large variation is confusing because there is little explanation
as to why there is so much of a dropoff in reaction time from 2013 to 2015 or the spike
from 2001 to 2003.  The variation from year to year is slightly concerning in the
sense that if there is a strong year effect then there is strong variation between
years.  This is not desirable and what this could possibly suggest is that
there is a lot of randomness in how athletes react or that there is at least one
other factor that affects reaction time. If the timing systems are succeptible to
external factors that affect how athletes react, then this is something that 
IAAF may want to consider when they select cities to host an important event 
such as the Track and Field World Championship. Additionally, Seiko may want to 
do more research on their timing systems to see if climate factors can impact
the precision of the accuracy of the equipment.

>>>>

2) and 3)
Other scholarly papers have discussed track and field reaction times but not in the
context of whether the minimum reaction time barrier is too low or whether timing
systems were accurate.  Outside the blog posts from LetsRun.com, there have not been
any published papers regarding Devon Allen's disqualification in particular.  This is not
necessarily surprising given that his disqualification occurred only last August.
Letsrun is a website that is part message board and part news.  The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as "Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and "The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the Letsrun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, statistics such as median reaction time, and a reference to a study
published in cooperation with IAAF in 2009 questioning the validity of the 0.1 
second reaction barrier.  The paper analyzed sprinters reaction times
and the biological processes that went into reacting to the starting gun and
concluded that athletes can react in as low as 0.08 seconds. It even recommended
to IAAF that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
This could provide a more accurate representation of reaction time, and thus the
possibility of a false disqualification would be lowered \citep{Komi}.
Despite the study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is confusing
as to why IAAF would commission a study only to disregard the results and not
make the necessary changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.

@@

4)
One of the first objectives of the paper is to try to model the data, which is
the reaction times of every non-disqualification since the 1999 World Championships. 
When determining the most effective model to use, several variations were
explored across the 4 different data sets: a linear model without a venue effect,
linear model with venue effect, a gamma model without a venue effect, a gamma model
with a venue effect, a gamma model without a venue effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed 
linear models with various distributions to be modeled\citep{Rpkg:lme4}.  The
model we use is a liner model with a fixed effect and random effect and follows
the gamma distribution. The fixed effect comes in the form of an intercept, and the
random effect will be called the venue effect and is reflective of how significant
the changes in the reaction time is for each venue, i.e. the venue effect is 
responsible for the variation between years.

Another objective of this paper is to compare data for athletes who have competed
at additional competitions to see how their reaction times compare from previous
competitions to this one. First we can look at the reaction times of athletes who competed at the
World Track and Field Championship in both 2019 and 2022 and compare the times,
and secondly we can look at how athletes performed compared to how they did at
their national/qualifiers meet.  We would expect that athletes of this caliber,
who are world-class athletes, to not improve significantly from one year to the
next.  In order to reduce any chances of athletes improving reaction time, we
can also look at data from qualifying meets.  Almost every athlete had to qualify 
by proving they are elite at the national level, usually by coming in at least
fourth. Thus, data was compiled on athletes who competed
at the 2022 World Track and Field Championships on what these athletes also ran 
at their country's national championship. If there was any argument that athletes
have improved their reaction time from 2019 to 2022, it should now be considered
nearly null, as the national competitions took place between June and July 2022 
and the World Championships were in August.  Thus, it should be expected that 
elite athletes such as these will not get reasonably faster in such a short time
span. The equipment used to measure reaction time cannot be guaranteed to be
the same from national to international competition, but the time span between
meets has been reduced considerably.

5)

@@

@@

\jy{Describe you data types; give details on sources; introduce notations for
  both types of data
  present exploratory plots.}
Something that is extremely important to note in the Data is the inconsistency of World Athletics 
in terms of how they penalize false starts.  From 2007-2009, the IAAF (the former name for World 
Athletics, and the governing body for the World Track and Field Champions) instituted a rule change
that allowed one warning false start before a sprinter was disqualified \citep{False-Start}  This 
essentially gave sprinters a second chance and reduced the penalty for a false start.  There were 
18 male and 7 female false starts at the 2007 World Championships, 18 male and 7 female starts at 
the 2009 World Championships. Starting in 2011, the rule was scrapped, and the old policy which 
automatically disqualifies runners who false start was reinstituted.  The effect? At the 2011 World
Championship there were 6 male disqualifications and 4 female disqualifications \citep{False-Start}. 
By returning to the harsher policy and cracking down on false starts, World Athletics had reduced the
number of false starts by two thirds. This was desirable for World Athletics as false starts can 
make already long track meets more tedious, both for athletes and viewers watching the 
television broadcast.  Since 2011, IAAF has instituted the one false start rule
\citep{Pilianidis}.

@@

Sorted By Model instead:
Linear Model with no venue effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Log Likelihood & Int effect est & Int se \\ [0.5ex] 
   \hline\hline
   All Finals & -472.7 & 238.3 & 0.1484 & 0.0019 \\
   \hline
   Older Finals & -450.7 & 227.4 & 0.1495 & 0.0019 \\ 
   \hline
   All Pooled & -1667 & 835.5 & 0.1529 & 0.0014 \\
   \hline
   Older Pooled & -1539 & 771.5 & 0.1544 & 0.0015 \\
   \hline
  \end{tabular}
  \end{center}

Linear model with venue effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Year se & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -468.7 & 0.0088 & 237.4 & 0.1481 & 0.0030 & 7.762e-05 & 2.457e-04 \\
   \hline
   Older Finals & -444.1 & 0.0075 & 225.0 & 0.1496 & 0.0028 & 5.691e-05 & 2.470e-04 \\ 
   \hline
   All Pooled & -1713 & 0.131 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\
   \hline
   Older Pooled & -1564 & 0.113 & 785.4 & 0.1562 & 0.0037 & 1.280e-04 & 6.273e-04 \\
   \hline
  \end{tabular}
  \end{center}

Gamma model with venue effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Year se & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -491.1 & 0.3388 & 248.5 & 6.8121 & 0.1895 & 0.1148 & 0.0112 \\
   \hline
   Older Finals & -465.1 & 0.2930 & 235.5 & 6.714 & 0.1674 & 0.0859 & 0.0109 \\ 
   \hline
   All Pooled & -1848 & 0.3612 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   Older Pooled & -1688 & 0.3145 & 846.8 & 6.426 & 0.1923 & 0.0989 & 0.0243 \\
   \hline
  \end{tabular}
  \end{center}

Gamma model with venue effect and link=log
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Year se & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -491.0 & 0.0504 & 248.5 & -1.914 & 0.0280 & 0.0025 & 0.0112 \\
   \hline
   Older Finals & -465.0 & 0.0439 & 235.5 & -1.901 & 0.0251 & 0.0019 & 0.0109 \\ 
   \hline
   All Pooled & -1848 & 0.0554 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\
   \hline
   Older Pooled & -1687 & 0.0495 & 846.6 & -1.853 & 0.0305 & 0.0024 & 0.0243 \\
   \hline
  \end{tabular}
  \end{center}

@@

Addition of 2022 on Gamma and log model on semifinal and finals data
\begin{center}
  \begin{tabular}{|c | c|}
    \hline
    Dataset & venue effect Standard Error \\
    1999-2019 & 0.0495 \\
    1999-2022 & 0.0554 \\
    \hline
  \end{tabular}
\end{center}

@@

Once the Gamma mixed effects model was chosen, it needed to be decided which combination
of effects to use.  It was possible to include a venue effect, a batch effect, or a year
and batch effect.  Density plots were created that simulated the probability of a mean
density less than 0.1, which is the current threshold for a reaction time
disqualification.  The density plots are shown below.  The batch effect is very
difficult to explain or interpret as it means that some heats of races were either
faster or slower and that everyone in the batch reacted similarly.  Although
it is possible that one athlete reacting faster could lead to others reacting
similarly, this does not make much practical sense as the difference between
heats that had fast and slow reaction times are all within 0.12 seconds of each
other.  It is much easier however to explain the venue effect.  The technology
that World Athletics has used has differed since 1999 and theoretically the
reaction time data from 2022 should be the best as the timing system should be
state of the art.  Any other weather or climate related factors such as humidity,
precipitation, elevation, etc. would all be negated when looking at the batch
effect, but are all factors that could potentially impact the venue effect.

@@

\section{Further Analysis}
, but that is not necessarily the results that were found.  The R package
"clusrank" was used to perform a clustered Wilcoxon rank sum test.  This was
chosen over a typical t test so as to avoid any assumptions about the distribution
of population means.  The clusrank package allows clusters to be formed, which
in this case are for each athlete.  The data containing each athlete's name,
nationality, and reaction times for the 2019 and 2022 World Championship can be
found in the csv file labeled "2019vs2022".  This data was transformed to "Clusrank2019vs2022"
so that it better fits the clusrank package.  Each athlete was given a non-meaningful
ID (An ID of 1 does not indicate that this athlete is better or worse than any
other athlete).  Additionally, the "Year" column is used to determine what year
the time was registered with "1" indicating 2019 and "2" indicating 2022.

@@

  Repeating the data collection steps and
again performing a clustered analysis for a Wilcoxon ranked sum test
results in two new z scores and p values.  The "ds" method produced a z score of
3.6069 and p value of 0.000155.  This result is highly significant and shows that
the times at the national stage of competition were significantly higher than
the reaction times at the World Championships even for an alpha level of 0.001.

@@

The figure below plots the average reaction time by year for the pooled men's data since
1999.  The circle in the bottom right is for 2022, which is significantly lower than the
data points for the previous three years, and is substially lower than the data points
for 2011 and 2013.  This graph supports the conclusion that the timing system from 2022
had some sort of error that caused the reaction times to be so much lower.  Assuming there
was not an issue with the system, the large variation is confusing because there is little explanation
as to why there is so much of a dropoff in reaction time from 2013 to 2015 or the spike
from 2001 to 2003.  The variation from year to year is slightly concerning in the
sense that if there is a strong venue effect then there is strong variation between
years.  This is not desirable and what this could possibly suggest is that
there is a lot of randomness in how athletes react or that there is at least one
other factor that affects reaction time. If the timing systems are succeptible to
external factors that affect how athletes react, then this is something that 
IAAF may want to consider when they select cities to host an important event 
such as the Track and Field World Championship. Additionally, Seiko may want to 
do more research on their timing systems to see if climate factors can impact
the precision of the accuracy of the equipment.
